---
title: "Devils Hole"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{devilshole}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette demonstrates the usage of `isogeochem` using a case study based on
[Bajnai et al. (2021)](https://doi.org/10.1029/2021GL093257).

#### The story:

We acquired a cave carbonate d18O time series, and we are interested in the variations in the d18O value of the groundwater. For this, we measured the clumped isotope values of ten samples. First, we will convert the D47 values to carbonate growth temperatures. Afterward, using these temperatures, we will convert the carbonate d18O time series to a groundwater d18O time series.

## Package setup

First, download and install `isogeochem` (you have to only do this once):

```{r, include = TRUE, message = FALSE, eval = TRUE}
if (!require("devtools")) install.packages("devtools")
devtools::install_github("davidbajnai/isogeochem")
```

To use `isogeochem`, load the package to the current R session (you have to do this every time you close and re-open R):

```{r, include = TRUE, message = FALSE, eval = TRUE}
library("isogeochem")
```

## Load data

Data can be loaded into R in many ways. For example, to load data from an excel file use the `openxlsx` package. For this vignette, however, lets specify variables manually:

```{r}
# D47(CDES90) values of Devils Hole carbonates 
DH_D47     = c(0.573, 0.575, 0.572, 0.581, 0.575, 0.575, 0.570, 0.574, 0.568, 0.575)
DH_D47_err = c(0.003, 0.007, 0.003, 0.005, 0.006, 0.003, 0.005, 0.005, 0.007, 0.005)
DH_D47_age = c(10.70, 36.00, 90.35, 122.75, 180.45, 236.65, 295.15, 355.65, 380.05, 496.65)
```

There are datasets available in `isogeochem`, which can be used simply by typing their name. For example, the `devilshole` dataset includes the original d18O composite time series from the Devils Hole caves.

```{r}
DH_age        = devilshole$age
DH_d18O_VSMOW = devilshole$d18O_VSMOW
DH_d18O_err   = devilshole$d18O_error
```

# Carbonate d18O versus age

Lets plot the carbonate d18O time series first:

```{r fig.height = 4, fig.width = 8}
# Convert d18O VSMOW values to the VPDB scale 
DH_d18O_VPDB = to_VPDB(DH_d18O_VSMOW)

# Calculate the errors
DH_d18O_VPDB_err1 = to_VPDB(DH_d18O_VSMOW + DH_d18O_err)
DH_d18O_VPDB_err2 = to_VPDB(DH_d18O_VSMOW - DH_d18O_err)

par(mfrow = c(1, 1), mar = c(4, 4.3, 0.5, 0.8))
plot(0, type="l", ylim = c(-17.5, -14.5), xlim = c(0, 570), xaxs="i", yaxs="i", xaxt = "n",
     ylab=expression(delta^18*"O"[calcite]*" (‰, VPDB)"), xlab="Age (ka)",
     lty=0, font=1, cex.lab=1, las = 1)

# Set up the x axis with ticks and labels
axis(1, at = seq(0, 570, by = 90))
axis(1, at = seq(0, 570, by = 10), labels=NA)

# Add the error interval the plot
polygon(c(DH_age, rev(DH_age)), c(DH_d18O_VPDB_err1, rev(DH_d18O_VPDB_err2)),
        col="wheat", border = NA)

# Add the carbonate d18O time series to the plot
lines(DH_age, DH_d18O_VPDB, lwd=2, col="gray10")
```

## Calculate growth temperatures from D47

```{r fig.height = 4, fig.width = 8}
# Convert D47 values into temperatures using the equation in Petersen et al. (2019)
DH_temp = temp_D47(DH_D47, eq = "Petersen19")
DH_temp_mean = mean(DH_temp)
DH_temp_mean

# Calculate the temperature error
DH_temp_err1 = temp_D47(DH_D47 + DH_D47_err, eq = "Petersen19")
DH_temp_err2 = temp_D47(DH_D47 - DH_D47_err, eq = "Petersen19")
DH_temp_err = mean((DH_temp_err1 - DH_temp_err2) / 2)
DH_temp_err

# Plot the results
par(mfrow = c(1, 1), mar = c(4, 4.3, 0.5, 0.8))
plot(0, type="l", ylim = c(31, 41), xlim = c(0, 570), xaxs="i", yaxs="i", xaxt = "n",
     ylab="Temperature (°C)", xlab="Age (ka)",
     lty=0, font=1, cex.lab=1, las = 1)

# Set up the x axis with ticks and labels
axis(1, at = seq(0, 570, by = 90))
axis(1, at = seq(0, 570, by = 10), labels=NA)

# Add mean error range
rect(0, DH_temp_mean-DH_temp_err,
     570, DH_temp_mean+DH_temp_err,
     col = "wheat", border = NA)

# Add error bars
segments(DH_D47_age, DH_temp_err1,
         DH_D47_age, DH_temp_err2, col = "gray50")

# Add points
points(DH_D47_age, DH_temp, col="gray10", pch=19)
```

## Groundwater d18O versus age

The calculated clumped isotope temperatures for the samples are indistinguishable, which allows us to calculate a mean temperature and use that to reconstruct the variations in groundwater.

```{r fig.height = 4, fig.width = 8}
# Calculate groundwater d18O values
DH_d18O_gw = d18O_H2O(DH_temp_mean, DH_d18O_VSMOW, min = "calcite", eq ="Coplen07")
DH_d18O_gw_err1 = d18O_H2O(DH_temp_mean + DH_temp_err, DH_d18O_VSMOW + DH_d18O_err,
                           min = "calcite", eq ="Coplen07")
DH_d18O_gw_err2 = d18O_H2O(DH_temp_mean - DH_temp_err, DH_d18O_VSMOW - DH_d18O_err,
                           min = "calcite", eq ="Coplen07")

# Plot the results
par(mfrow = c(1, 1), mar = c(4.1, 4.4, 0.5, 0.8))
plot(0, type="l", ylim = c(-15, -11), xlim = c(0, 570), xaxs="i", yaxs="i", xaxt = "n",
     ylab=expression(delta^18*"O"[groundwater]*" (‰, VSMOW)"), xlab="Age (ka)",
     lty=0, font=1, cex.lab=1, las = 1)

# Set up the x axis with ticks and labels
axis(1, at = seq(0, 570, by = 90))
axis(1, at = seq(0, 570, by = 10), labels=NA)

# Add the error interval the plot
polygon(c(DH_age, rev(DH_age)), c(DH_d18O_gw_err1, rev(DH_d18O_gw_err2)),
        col="wheat", border = NA)

# Add the carbonate d18O time series to the plot
lines(DH_age, DH_d18O_gw, lwd=2, col="gray10")
```
