---
title: "Kim and O'Neil (1997) recalculation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Kim and O'Neil (1997) recalculation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
#> Save the user's options and parameters
oldopt = options()
oldpar = par()
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  eval = TRUE,
  fig.path = "figures/kim_oneil_1997-",
  fig.height = 5,
  fig.width = 5,
  dpi = 150,
  out.width = "55%"
)
knitr::opts_knit$set(global.par = TRUE)
r = getOption("repos")
r["CRAN"] = "https://stat.ethz.ch/CRAN/"
options(repos = r)
```

```{r, include = FALSE}
#> Set new parameters for this vignette.
par(mfrow = c(1, 1), mar = c(4.5, 4.5, 0.3, 0.3))
```

## Introduction

This vignette provides a detailed description of the reprocessing of the <sup>18</sup>O/<sup>16</sup>O fractionation factor between calcite and water provided by [Kim and O'Neil (1997)](https://doi.org/10.1016/S0016-7037(97)00169-5):

**10<sup>3</sup>ln&alpha; = 18.03 x 1000 / T - 32.42**

To calculate calcite *δ*<sup>18</sup>O values from the *δ*<sup>18</sup>O values of CO<sub>2</sub> produced from acid digestion, Kim and O'Neil (1997) used an <sup>18</sup>O/<sup>16</sup>O acid fractionation factor (AFF) of 1.01050 at 25 °C. However, the current IUPAC recommendation for an AFF at 25°C is 1.010254 (see Kim et al. 2007 and 2015). The difference between calcite *δ*<sup>18</sup>O values calculated using the two acid fractionation factors is ca. 0.24‰.

To be able to apply the Kim and O'Neil (1997) equation to calcite *δ*<sup>18</sup>O data produced with the IUPAC-recommended acid fractionation factors and to compare the Kim and O'Neil (1997) equation with more recent oxygen isotope paleothermometry equations, such as the Daëron et al. (2019) equation, the original data has to be reprocessed.

## Package setup

Download, install, and load the `isogeochem` package:

```{r, include = TRUE, message = FALSE, eval = TRUE}
if (!require("isogeochem")) install.packages("isogeochem")
library("isogeochem")
```

## Load data

The data is from Table 1 in Kim and O'Neil (1997). All *δ*<sup>18</sup>O values are expressed on the VSMOW scale.

```{r, include = TRUE, message = FALSE, eval = TRUE}
TinC = c(10, 10, 25, 25, 25, 25, 40, 40, 40)
TinK = 1000 / (TinC + 273.15)
d18O_H2O = c(-8.12, -8.23, -8.30, -8.25, -8.12, -8.23, -8.20, -8.12, -8.23)
d18O_calcite = c(23.47, 23.21, 19.73, 20.23, 20.00, 20.03, 17.06, 17.24, 17.01)
```

## The original equation

Lets reproduce the slope and intercept of the original equation.

```{r, include = TRUE, message = FALSE, eval = TRUE}
# Calculate the fractionation factor between calcite and water
a18_calcite_H2O = a_A_B(A = d18O_calcite, B = d18O_H2O)

# Calculate the 1000ln alpha values, abbreviated here as "elena"
# Kim and O'Neil (1997) used values rounded to two decimals
elena_orig = round(1000 * log(a18_calcite_H2O), 2)

# Fit a linear regression on the values 
lm_orig = lm(elena_orig ~ TinK)
slope_orig = round(as.numeric(coef(lm_orig)["TinK"]), 2)
intercept_orig = round(as.numeric(coef(lm_orig)["(Intercept)"]), 2)

# The original equation:
slope_orig
intercept_orig
```

## Reprocess the calcite *δ*<sup>18</sup>O values

```{r}
# Convert d18O_calcite to d18O_CO2acid using the "old" AFF
d18O_CO2acid = A_from_a(a = 1.01050, B = d18O_calcite)

# Convert d18O_CO2acid to d18O_calcite using the "new" AFF
AFF_new = a18_CO2acid_c(25, "calcite")
d18O_calcite_newAFF = B_from_a(a = AFF_new, d18O_CO2acid)
```

## Determine the slope and intercept using the new *δ*<sup>18</sup>O values

```{r}
# Calculate the new alpha and 1000ln alpha values
a18_calcite_H2O_new = a_A_B(A = d18O_calcite_newAFF, B = d18O_H2O)
elena_new = 1000 * log(a18_calcite_H2O_new)

# Calculate new slope and intercept
lm_new = lm(elena_new ~ TinK)
slope_new = round(as.numeric(coef(lm_new)["TinK"]), 2)
intercept_new = round(as.numeric(coef(lm_new)["(Intercept)"]), 2)
slope_new
intercept_new
```

Consequently, the reprocessed equation is:

**10<sup>3</sup> ln &alpha; = 18.04 x 1000 / T - 32.18**

## Visualize the 10<sup>3</sup>ln&alpha; vs. temperature relationships

```{r, label = "Figure1"}
if (!require("viridisLite")) install.packages("viridisLite")

plot(0, type = "l", las = 1,
     ylim = c(28, 33),
     xlim = c(10, 30),
     ylab = expression("1000 ln " * alpha[calcite / water] ^ 18 * " (‰)"),
     xlab = "Temperature (°C)")

temp = seq(10, 30, 1)
cols = viridisLite::viridis(6, option = "D")

lines(temp, 1000 * log(a18_c_H2O(temp, "calcite", "Daeron19")),
      col = cols[1], lwd = 2)
lines(temp, 1000 * log(a18_c_H2O(temp, "calcite", "Watkins13")),
      col = cols[2], lwd = 2)
lines(temp, 1000 * log(a18_c_H2O(temp, "calcite", "Tremaine11")),
      col = cols[3], lwd = 2)
lines(temp, 1000 * log(a18_c_H2O(temp, "calcite", "ONeil69")),
      col = cols[4], lwd = 2)
lines(temp, 1000 * log(a18_c_H2O(temp, "calcite", "KO97")),
      col = cols[5], lwd = 2)
lines(temp, 1000 * log(a18_c_H2O(temp, "calcite", "KO97-orig")),
      col = cols[6], lwd = 2, lty = 2)

legend("topright", bty = "n", adj = c(0, NA), 
       lty = c(1, 1, 1, 1, 1, 3),
       lwd = c(2, 2, 2, 2, 3, 3),  
       col = cols,
       legend = c("Daeron19",
                  "Watkins13",
                  "Tremaine11",
                  "ONeil69",
                  "KO97",
                  "KO97-orig"))
```

```{r, include = FALSE}
#> Reset to the user's options and parameters.
par(oldpar)
options(oldopt)
```
